Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-12-22 threshold/arxiv.tex
Added arxiv version

%% Instructions for processing this document into a PDF document
%%
%% FIRST NOTE, you need the 'xtable' and 'biclust' packages installedi in R.
%% Some of the examples will come from data sets in this package.
%% You can install it from the R console with, for example:
%%
%% install.packages("biclust",repos="http://cran.case.edu")
%% install.packages("xtable",repos="http://cran.case.edu")
%%
%% Now you can prepare the PDF of the paper with any of:
%%
%% From a shell:
%% R CMD Sweave --pdf draft.Rnw
%%
%% From an R prompt:
%% Sweave("draft.Rnw")
%% tools::texi2pdf("draft.tex")
%%
%% From RStudio:
%% click on the little "Compile PDF" button
%%
%% Note that you should probably run the above a few times because
%% LaTeX generally takes a few runs to get reference numbers right.

\documentclass[article]{jss}
%% need no \usepackage{Sweave.sty}

\usepackage{amsmath}
\usepackage{hyperref}

@@

\title{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
\author{
James Baglama\\
University of Rhode Island
\And
Michael Kane\\
Yale University
\AND
Bryan Lewis\\
Paradigm4, Inc.\\
\And
Alex Poliakov\\
Paradigm4, Inc.
}
\Plainauthor{James Baglama, Michael Kane, Bryan Lewis, Alex Poliakov}
\Plaintitle{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
\Shorttitle{Efficient Thresholded Correlation}

\Abstract{
Efficiently computing a subset of a correlation matrix consisting of values
above a specified threshold is important to many practical applications.
Real-world problems in genomics, machine learning, finance other applications
can produce correlation matrices too large to explicitly form and tractably
compute. Often, only values corresponding to highly-correlated vectors are of
interest, and those values typically make up a small fraction of the overall
correlation matrix. We present a method based on the singular value
decomposition (SVD) and its relationship to the data covariance structure that
can efficiently compute thresholded subsets of very large correlation matrices.
}
\Keywords{thresholded correlation, SVD, \proglang{R}}
\Plainkeywords{thresholded correlation, SVD, R}
\Address{
James Baglama, Chair of Mathematics\\
Department of Mathematics\\
University of Rhode Island\\
Kingston, Rhode Island 02881\\
E-mail: \email{jbaglama@math.uri.edu}\\
URL: \url{http://www.math.uri.edu/~jbaglama}\\
\\
Michael Kane\\
Yale University\\
300 Geroge Street\\
New Haven, Connecticut 06511\\
E-mail: \email{michael.kane@yale.edu}\\
\\
Bryan Lewis\\
Paradigm4, Inc.\\
281 Winter Street, Suite 360\\
Waltham, MA 02451\\
E-mail: \email{blewis@illposed.net}\\
URL: \url{http://illposed.net}\\
\\
Alex Poliakov\\
Paradigm4, Inc.\\
281 Winter Street, Suite 360\\
Waltham, MA 02451\\
E-mail: \email{apoliakov@paradigm4.com}\\
}

\begin{document}

@@

Finding highly-correlated pairs among a large set of vectors is an important
part of many applications. For instance, subsets of highly-correlated vectors
of gene expression values may be used in the discovery of networks of genes
relevant to particular biological processes \citep{genomics}.  Identification of
highly-correlated pairs may be used in feature selection algorithms for machine
learning applications \citep{ml1, ml2} and are important to time series and
image processing applications \citep{timeseries, svd-similarity}.

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasible for high-dimensional data in modern genomics and other applications.
Considerable attention has been devoted to \emph{pruning
methods} that cheaply prune away all but the most highly-correlated pairs of
vectors.  For example the TAPER and related algorithms of Xiong et al. develop
pruning rules based on the frequency of mutually co-occurring observations (rows)
among vector pairs; see \citep{prune1, prune2} and the references therein.
Related methods approximate a set containing the most highly-correlated
pairs using hashing algorithms \citep{prune3}. A number of other methods arrive
at approximate sets of the most highly-correlated pairs using
dimensionality-reduction techniques like the discrete Fourier
transform \citep{timeseries}. The method of Wu, et al. \citep{svd-similarity},
conceptually very similar to our approach, uses the singular value
decomposition (SVD) to find all pairs of close vectors with respect to a
distance metric (instead of correlation).  Our method finds all of the most
highly-correlated vector pairs with respect to a given threshold by pruning
along an intuitive path of decreasing variance defined by the data and computed
by a truncated SVD.

@@

Most of the floating point operations occur in step 4 of Algorithm \ref{prune}.
Fortunately the $\ell$ evaluations in step 4 are independent of each other and
can easily be computed in parallel. See the reference \proglang{R}
implementation \citep{sup} for an example.

@@

The augmented implicitly restarted Lanczos bidiagonalization algorithm\break(IRLBA)
of Baglama and Reichel \citep{irlba} efficiently computes truncated singular value
decompositions of large dense or sparse matrices. Reference implementations are
available for \proglang{R} \citep{irlbar}, \proglang{Matlab} \citep{irlbam},
and \proglang{Python} \citep{irlbap}.

We can relax the column mean and scale assumptions without introducing much
additional computational or storage overhead as follows. Assume $A$ is an
$m\times n$ real-valued matrix without any constant-valued columns. Let
$z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
of $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
be the vector of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$
a diagonal matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
matrix with zero column means and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
\]
The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
products. The main idea behind efficient application of IRLB to correlation
problems replaces matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ in
the iterations, implicitly working with a scaled and centered matrix without
forming it.  This comes at the cost of storing two additional length $n$ vectors and
at most computing two additional vector inner products per matrix-vector product.
The \proglang{R} implementation \citep{irlbar} of IRLBA includes arguments for centering and
scaling the input matrix.

@@

We evaluated the algorithm using public gene expression and DNA methylation
datasets from the \proglang{R} {\tt biclust} \citep{biclust} package and from
the Cancer Genome Atlas \citep{gdac}, referred to as TCGA.  Most tests were
performed on a desktop PC with a single AMD A10-7850K 3.7$\,$GHz Athlon
quad-core CPU and 16$\,$GB 1333$\,$MHz unbuffered RAM running the Ubuntu 15.10
GNU/Linux OS. Tests were performed with 64-bit \proglang{R} version 3.2.3 using
double-precision arithmetic and the OpenBLAS library version 0.2.14-1ubuntu1
(based on Goto's BLAS version 1.13) with {\tt OMP\_NUM\_THREADS=1}. The last
experiment includes a test run on a Linux cluster described in that section.

@@

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} \proglang{R} package \citep{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as an $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95. The choice of projection dimension $p$ affects
the efficiency of the algorithm. For $p=2$ the number of candidate vector pairs
generated in step 4 of Algorithm \ref{prune} is over 3 million. When $p$ is increased
to $10$, the number of candidate vector pairs drops to less than 13,000.
The reference \proglang{R} implementation includes options for
tuning $p$ and relies on the restart ability of the IRLBA to pick up where it left off and
inexpensively extend an existing truncated SVD. The following example uses $p=10$.
\begin{small}
\begin{Schunk}
\begin{Sinput}
> threshold = 0.95
> # Test system values:
> t1=3.13
> t2=1.75
> m1=59.60
> m2=1075.10
> if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
+ {
+ library(tcor)
+ library(doMC)     # local parallel computation
+ library(biclust)  # for the BicatYeast data
+ data(EisenYeast)  # in gene by sample orientation
+ A = t(EisenYeast) # to correlate genes to genes
+ # Register a four-core multicore parallel backend
+ registerDoMC(4)
+ 
+ # Compute using tcor():
+ m1  = sum(gc()[,2])  # Memory in megabytes currently in use
+ t1  = proc.time()
+ tx  = tcor(A, t=threshold, p=10)
+ t1  = (proc.time() - t1)[3] # Time in seconds
+ m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test
+ 
+ # Fast brute force full correlation:
+ m2  = sum(gc()[,2])
+ t2  = proc.time()
+ mu = colMeans(A)
+ s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
+ cs = scale(A, center=mu, scale=s)  # scale and center the data
+ cx = crossprod(cs)                 # full correlation matrix
+ cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
+ pairs = which(cx >= threshold, arr.ind=TRUE)
+ t2  = (proc.time() - t2)[3]
+ m2  = sum(gc()[,6]) - m2
+ }
\end{Sinput}
\end{Schunk}
\end{small}
Brute-force computation outperforms {\tt tcor()} in this small first example in
speed but consumes substantially more memory.  Each algorithm identifies the
same set of 125 gene expression pairs that exceed the specified correlation
threshold of 0.95, with comparative timing and memory use shown in the
following table.

@@

Performance benefits from the {\tt tcor()} function become apparent with more
data.  The following example uses calculated beta values from the Illumina
Human Methylation450 BeadChip of 80 Adenoid Cystic Carcinoma (ACC) tumor
samples obtained from the Broad GDAC Cancer Genome Atlas dashboard \citep{gdac}.
See the \proglang{R} package implementation \citep{sup} for examples of
efficiently converting the downloaded sample methylation data into an R matrix.
We chose these data for their public availability as a general example of
computing thresholded correlation among large numbers of vectors,
rather than for any specific biological meaning.

@@

No code change was required to run in parallel on the cluster. The reference \proglang{R}
{\tt tcor()} implementation uses \proglang{R}'s {\tt foreach} \citep{foreach} framework to
run on arbitrary parallel configurations (``back ends''). We simply registered
a {\tt doRedis} \citep{doredis} parallel back end with 16 \proglang{R} workers per computer
(64 total \proglang{R} worker processes) prior to invoking {\tt tcor()}.

@@

The idea of evaluating threshold conditions by projection into low dimensional
subspaces applies to distance metrics similarly to correlation, see
\citep{svd-similarity} for a related approach. Lemma \ref{sumlemma} decomposes
Euclidean distance between two vectors in terms of singular basis vectors and
Algorithm \ref{prune} can be applied directly to the problem of finding all
pairs of vectors that meet a thresholded Euclidean distance with small changes
in steps 3, 4, and 7.  By equivalence of norms, thresholded Euclidean distance
can be extended to other distance metrics. The reference \proglang{R}
implementation \citep{sup} includes functions for thresholded distance and
correlation.

The ideas presented here can be adapted to efficiently find the top N most
correlated vector pairs for a given constant N similarly to the work of
Xiong et al. \citep{prune2}.

@@

\bibliography{bibliography}{}

>>>>

\documentclass{article}

\usepackage{Sweave}
\usepackage{cite}
\usepackage{authblk}

\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{graphicx}

\let\proglang=\textsf

@@

\title{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
%%\author{
%%James Baglama\\
%%University of Rhode Island
%%\And
%%Michael Kane\\
%%Yale University
%%\AND
%%Bryan Lewis\\
%%Paradigm4, Inc.\\
%%\And
%%Alex Poliakov\\
%%Paradigm4, Inc.
%%}
%%\Plainauthor{James Baglama, Michael Kane, Bryan Lewis, Alex Poliakov}
%%\Plaintitle{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
%%\Shorttitle{Efficient Thresholded Correlation}
\author[1]{James Baglama \thanks{jbaglama@math.uri.edu}}
\author[2]{Michael Kane \thanks{michael.kane@yale.edu}}
\author[3]{Bryan Lewis \thanks{blewis@illposed.net}}
\author[3]{Alex Poliakov \thanks{apoliakov@paradigm4.com}}
\affil[1]{
University of Rhode Island
}
\affil[2]{
Yale University
}
\affil[3]{
Paradigm4, Inc.
}
\renewcommand\Authands{ and }
%\Abstract{
\begin{document}
\maketitle
\begin{abstract}
Efficiently computing a subset of a correlation matrix consisting of values
above a specified threshold is important to many practical applications.
Real-world problems in genomics, machine learning, finance other applications
can produce correlation matrices too large to explicitly form and tractably
compute. Often, only values corresponding to highly-correlated vectors are of
interest, and those values typically make up a small fraction of the overall
correlation matrix. We present a method based on the singular value
decomposition (SVD) and its relationship to the data covariance structure that
can efficiently compute thresholded subsets of very large correlation matrices.
\end{abstract}
%%\Keywords{thresholded correlation, SVD, \proglang{R}}
%%\Plainkeywords{thresholded correlation, SVD, R}
%%\Address{
%%James Baglama,
%%Chair of Mathematics\\
%%Department of Mathematics\\
%%University of Rhode Island\\
%%Kingston, Rhode Island 02881\\
%%E-mail: \email{jbaglama@math.uri.edu}\\
%%URL: \url{http://www.math.uri.edu/~jbaglama}\\
%%\\
%%Michael Kane\\
%%Yale University\\
%%300 Geroge Street\\
%%New Haven, Connecticut 06511\\
%%E-mail: \email{michael.kane@yale.edu}\\
%%\\
%%Bryan Lewis\\
%%Paradigm4, Inc.\\
%%281 Winter Street, Suite 360\\
%%Waltham, MA 02451\\
%%E-mail: \email{blewis@illposed.net}\\
%%URL: \url{http://illposed.net}\\
%%\\
%%Alex Poliakov\\
%%Paradigm4, Inc.\\
%%281 Winter Street, Suite 360\\
%%Waltham, MA 02451\\
%%E-mail: \email{apoliakov@paradigm4.com}\\
%%}

@@

Finding highly-correlated pairs among a large set of vectors is an important
part of many applications. For instance, subsets of highly-correlated vectors
of gene expression values may be used in the discovery of networks of genes
relevant to particular biological processes \cite{genomics}.  Identification of
highly-correlated pairs may be used in feature selection algorithms for machine
learning applications \cite{ml1, ml2} and are important to time series and
image processing applications \cite{timeseries, svd-similarity}.

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasible for high-dimensional data in modern genomics and other applications.
Considerable attention has been devoted to \emph{pruning
methods} that cheaply prune away all but the most highly-correlated pairs of
vectors.  For example the TAPER and related algorithms of Xiong et al. develop
pruning rules based on the frequency of mutually co-occurring observations (rows)
among vector pairs; see \cite{prune1, prune2} and the references therein.
Related methods approximate a set containing the most highly-correlated
pairs using hashing algorithms \cite{prune3}. A number of other methods arrive
at approximate sets of the most highly-correlated pairs using
dimensionality-reduction techniques like the discrete Fourier
transform \cite{timeseries}. The method of Wu, et al. \cite{svd-similarity},
conceptually very similar to our approach, uses the singular value
decomposition (SVD) to find all pairs of close vectors with respect to a
distance metric (instead of correlation).  Our method finds all of the most
highly-correlated vector pairs with respect to a given threshold by pruning
along an intuitive path of decreasing variance defined by the data and computed
by a truncated SVD.

@@

Most of the floating point operations occur in step 4 of Algorithm \ref{prune}.
Fortunately the $\ell$ evaluations in step 4 are independent of each other and
can easily be computed in parallel. See the reference \proglang{R}
implementation \cite{sup} for an example.

@@

The augmented implicitly restarted Lanczos bidiagonalization algorithm\break(IRLBA)
of Baglama and Reichel \cite{irlba} efficiently computes truncated singular value
decompositions of large dense or sparse matrices. Reference implementations are
available for \proglang{R} \cite{irlbar}, \proglang{Matlab} \cite{irlbam},
and \proglang{Python} \cite{irlbap}.

We can relax the column mean and scale assumptions without introducing much
additional computational or storage overhead as follows. Assume $A$ is an
$m\times n$ real-valued matrix without any constant-valued columns. Let
$z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
of $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
be the vector of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$
a diagonal matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
matrix with zero column means and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
\]
The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
products. The main idea behind efficient application of IRLB to correlation
problems replaces matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ in
the iterations, implicitly working with a scaled and centered matrix without
forming it.  This comes at the cost of storing two additional length $n$ vectors and
at most computing two additional vector inner products per matrix-vector product.
The \proglang{R} implementation \cite{irlbar} of IRLBA includes arguments for centering and
scaling the input matrix.

@@

We evaluated the algorithm using public gene expression and DNA methylation
datasets from the \proglang{R} {\tt biclust} \cite{biclust} package and from
the Cancer Genome Atlas \cite{gdac}, referred to as TCGA.  Most tests were
performed on a desktop PC with a single AMD A10-7850K 3.7$\,$GHz Athlon
quad-core CPU and 16$\,$GB 1333$\,$MHz unbuffered RAM running the Ubuntu 15.10
GNU/Linux OS. Tests were performed with 64-bit \proglang{R} version 3.2.3 using
double-precision arithmetic and the OpenBLAS library version 0.2.14-1ubuntu1
(based on Goto's BLAS version 1.13) with {\tt OMP\_NUM\_THREADS=1}. The last
experiment includes a test run on a Linux cluster described in that section.

@@

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} \proglang{R} package \cite{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as an $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95. The choice of projection dimension $p$ affects
the efficiency of the algorithm. For $p=2$ the number of candidate vector pairs
generated in step 4 of Algorithm \ref{prune} is over 3 million. When $p$ is increased
to $10$, the number of candidate vector pairs drops to less than 13,000.
The reference \proglang{R} implementation includes options for
tuning $p$ and relies on the restart ability of the IRLBA to pick up where it left off and
inexpensively extend an existing truncated SVD. The following example uses $p=10$.
\begin{small}
\begin{Schunk}
\begin{Sinput}
> threshold = 0.95
> # Test system values:
> t1=3.13
> t2=1.75
> m1=59.60
> m2=1075.10
> if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
+ {
+ library(tcor)
+ library(doMC)     # local parallel computation
+ library(biclust)  # for the BicatYeast data
+ data(EisenYeast)  # in gene by sample orientation
+ A = t(EisenYeast) # to correlate genes to genes
+ # Register a four-core multicore parallel backend
+ registerDoMC(4)
+ 
+ # Compute using tcor():
+ m1  = sum(gc()[,2])  # Memory in megabytes currently in use
+ t1  = proc.time()
+ tx  = tcor(A, t=threshold, p=10)
+ t1  = (proc.time() - t1)[3] # Time in seconds
+ m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test
+ 
+ # Fast brute force full correlation:
+ m2  = sum(gc()[,2])
+ t2  = proc.time()
+ mu = colMeans(A)
+ s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
+ cs = scale(A, center=mu, scale=s)  # scale and center the data
+ cx = crossprod(cs)                 # full correlation matrix
+ cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
+ pairs = which(cx >= threshold, arr.ind=TRUE)
+ t2  = (proc.time() - t2)[3]
+ m2  = sum(gc()[,6]) - m2
+ }
\end{Sinput}
\end{Schunk}
\end{small}
Brute-force computation outperforms {\tt tcor()} in this small first example in
speed but consumes substantially more memory.  Each algorithm identifies the
same set of 125 gene expression pairs that exceed the specified correlation
threshold of 0.95, with comparative timing and memory use shown in the
following table.

@@

Performance benefits from the {\tt tcor()} function become apparent with more
data.  The following example uses calculated beta values from the Illumina
Human Methylation450 BeadChip of 80 Adenoid Cystic Carcinoma (ACC) tumor
samples obtained from the Broad GDAC Cancer Genome Atlas dashboard \cite{gdac}.
See the \proglang{R} package implementation \cite{sup} for examples of
efficiently converting the downloaded sample methylation data into an R matrix.
We chose these data for their public availability as a general example of
computing thresholded correlation among large numbers of vectors,
rather than for any specific biological meaning.

@@

No code change was required to run in parallel on the cluster. The reference \proglang{R}
{\tt tcor()} implementation uses \proglang{R}'s {\tt foreach} \cite{foreach} framework to
run on arbitrary parallel configurations (``back ends''). We simply registered
a {\tt doRedis} \cite{doredis} parallel back end with 16 \proglang{R} workers per computer
(64 total \proglang{R} worker processes) prior to invoking {\tt tcor()}.

@@

The idea of evaluating threshold conditions by projection into low dimensional
subspaces applies to distance metrics similarly to correlation, see
\cite{svd-similarity} for a related approach. Lemma \ref{sumlemma} decomposes
Euclidean distance between two vectors in terms of singular basis vectors and
Algorithm \ref{prune} can be applied directly to the problem of finding all
pairs of vectors that meet a thresholded Euclidean distance with small changes
in steps 3, 4, and 7.  By equivalence of norms, thresholded Euclidean distance
can be extended to other distance metrics. The reference \proglang{R}
implementation \cite{sup} includes functions for thresholded distance and
correlation.

The ideas presented here can be adapted to efficiently find the top N most
correlated vector pairs for a given constant N similarly to the work of
Xiong et al. \cite{prune2}.

@@

\bibliography{bibliography}{}
\bibliographystyle{plain}
